Supplementary material for: non-coding RNAs of bird genomes

### 2014-05-23 paper/supplementary-material.tex
Spell check.

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird dataset.
Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 loci of the ancestral
vertebrate genome. We investigated these losses further.

@@

The turkey genome contains a high number (190) of microRNAs so far
found only in chicken, which account for the higher number of
annotated sequences in this genome compared with other birds. This is
consistent with its phylogenetic position as the closest chicken
relative among the examined birds. However, 101 chicken microRNAs
have no homolog in the turkey or other bird genomes, suggesting that
these genes are chicken-specific. This is consistent with previous
reports of large number of speciesspecific microRNAs in all animals,
and supports the view of fast microRNA turnover during animal
evolution.

@@

The simplest case are the selenocysteine tRNAs. Here, tRNAscan is
tuned for specificity and thus misses several occurrances that are
easily found by \texttt{blastn} with $E\le 10^{-30}$. In some cases
the sequences appear degraded at the ends, which may be explained
e.g.\ by low sequence quality at the very ends of contigs or
scaffolds. A \texttt{blastn} search also readily retrieves additional
RNase P and RNAse MRP RNAs, capturing only the best conserved
regions. In many cases these additional candidates are incomplete or
contain undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Vault RNA homology
searches with \texttt{blastn} remained unsuccessful, therefore we
constructed a sauropsid-specific CM.  In addition to the hits
identified by the Rfam model we obtained three additional
homologs. Vault RNAs, with a size of about 100 nt, exhibit conserved
sequence patterns only at their ends, with essentially unconstrained
sequence in the central part. Their identification is one of the
well-known and difficult problems for homology search
\cite{Stadler:09b,Kolbe:2009}.

@@

\begin{figure}[ht]
  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
    the left.}\label{fig:3}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/tRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf tRNA} families.}\label{fig:5}
\end{figure}

>>>>

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird data set.
Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 loci of the ancestral
vertebrate genome. We investigated these losses further.

@@

The turkey genome contains a high number (190) of microRNAs so far
found only in chicken, which account for the higher number of
annotated sequences in this genome compared with other birds. This is
consistent with its phylogenetic position as the closest chicken
relative among the examined birds. However, 101 chicken microRNAs
have no homolog in the turkey or other bird genomes, suggesting that
these genes are chicken-specific. This is consistent with previous
reports of large number of species specific microRNAs in all animals,
and supports the view of fast microRNA turnover during animal
evolution.

@@

The simplest case are the selenocysteine tRNAs. Here, tRNAscan is
tuned for specificity and thus misses several occurrences that are
easily found by \texttt{blastn} with $E\le 10^{-30}$. In some cases
the sequences appear degraded at the ends, which may be explained
e.g.\ by low sequence quality at the very ends of contigs or
scaffolds. A \texttt{blastn} search also readily retrieves additional
RNase P and RNAse MRP RNAs, capturing only the best conserved
regions. In many cases these additional candidates are incomplete or
contain undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Vault RNA homology
searches with \texttt{blastn} remained unsuccessful, therefore we
constructed a sauropsid-specific CM.  In addition to the hits
identified by the Rfam model we obtained three additional
homologs. Vault RNAs, with a size of about 100 nt, exhibit conserved
sequence patterns only at their ends, with essentially unconstrained
sequence in the central part. Their identification is one of the
well-known and difficult problems for homology search
\cite{Stadler:09b,Kolbe:2009}.

@@

\begin{figure}[ht]
  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
  \caption[]{Heatmaps showing the presence/absence and approximate
    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
    the left.}\label{fig:3}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/tRNA.pdf}
  \caption[]{Heatmaps showing the presence/absence and approximate
    copy number of {\bf tRNA} families.}\label{fig:5}
\end{figure}
